---
title: "Exploring the U.S. National Oceanic and Atmospheric Administrations' (NOAA) storm database"
author: "Cesar Augusto JImenez Sanchez"
date: "11/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

- Storm Data.

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

```{r import-data, echo = FALSE}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!dir.exists("./data")){
      dir.create("data")
      print("Data directory was created.")
} else if(!file.exists("./data/storm_data.csv.bz2")) {
      download.file(url, "./data/storm_data.csv.bz2")
      print("Storm data was downloaded successfully.")
}

storm <- read_csv("./data/storm_data.csv.bz2")
```

## Questions

Your data analysis must address the following questions:

1. Across the United States, which types of events (as indicated in the `EVTYPE` variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.

## Requirements
For this assignment you will need some specific tools

- RStudio: You will need RStudio to publish your completed analysis document to RPubs. You can also use RStudio to edit/write your analysis.
- knitr: You will need the knitr package in order to compile your R Markdown document and convert it to HTML

## Document Layout
- Language: Your document should be written in English.
- Title: Your document should have a title that briefly summarizes your data analysis
- Synopsis: Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.
- There should be a section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the \color{red}{\verb|cache = TRUE|}cache = TRUE option for certain code chunks.
- There should be a section titled Results in which your results are presented.
- You may have other sections in your analysis, but Data Processing and Results are required.
- The analysis document must have at least one figure containing a plot.
- Your analysis must have no more than three figures. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.
- You must show all your code for the work in your analysis document. This may make the document a bit verbose, but that is okay. In general, you should ensure that `echo = TRUE` for every code chunk (this is the default setting in knitr).